Parsing Postfix log files

### 2009-03-02 doc/logparser-architecture.tex
Rewrote some of the actions section to deal with XXX items. Wrote the cascaded parsing section. Added script to demonstrate cascaded parsing.

Each action is a separate procedure written to process a particular
category of input, e.g.\ rejecting a delivery attempt.  There may be many
input variants within one input category; in general each action will
handle one category of inputs, with each rule recognising one input
variant.  XXX EXPAND AND CLARIFY\@; THE MESSAGE IS HERE, IT JUST NEEDS TO
BE CLEARER\@.  It is anticipated that parsers based on this architecture
will have a high ratio of rules to actions, with the aim of having simple
rules and clear distinctions between the inputs parsed by different rules.
An action may process different input variants in slightly different ways,
but large changes in processing indicate the need for a new action and a
new category of input; if an action becomes unnecessarily complicated it
starts to turn into a monolithic parser, with too much logic contained in a
single procedure.

The ability to easily add special purpose actions to deal with difficulties
and new requirements that are discovered during parser development is one
of the strengths of this architecture.  XXX REWRITE, EXPAND, AND CLARIFY UP
TO THE NEXT XXX\@.  Instead of writing a single monolithic function that
must be modified to support new behaviour, with all the attendant risks of
adversely affecting the existing parser, when a new requirement arises an
independent action can be written to satisfy it.  Sometimes the new action
will require the cooperation of other actions, e.g.\ to set or check a
flag, or save some extra data.  There is a possibility of introducing
failure when modifying existing actions in this way, but the modifications
will be smaller and occur less frequently than with a monolithic
architecture, thus failures will be less likely to occur and will be easier
to test for and diagnose.  XXX FINISH REWRITING\@.  The architecture can be
implemented in an object oriented style, allowing sub-classes to extend or
override actions in addition to adding new actions; because each action is
an independent procedure, the sub-class need only modify the action it is
overriding, and can inherit the others rather than reproducing large chunks
of functionality.

During development of the Postfix log parser it became apparent that in
addition to the obvious variety in log~lines there were
many complications to be overcome.  Some were the result of deficiencies in
Postfix's logging (some of which were rectified by later versions of
Postfix, e.g.\ identifying bounce
notifications~\sectionref{identifying-bounce-notifications}); others were
due to the vagaries of process scheduling, client behaviour, and
administrative actions.  All were successfully accommodated in the Postfix
log parser: adding new actions was enough to overcome several of the
complications; others required modifications to a single existing action to
work around the difficulties; the remainder were resolved by adapting
existing actions to cooperate and exchange extra data (via the framework),
changing their behaviour as appropriate based on that extra data.  Every
architecture should aim to make the hard things possible as well as making
the easy things easy; the successful implementation of the Postfix log
parser demonstrates that this architecture achieves that aim.

Actions are not context-sensitive, but they can consult the results (or
lack of results) of previous actions during execution, providing some
context sensitivity.  XXX EXPAND\@.

@@

Actions may return a modified input line that will be parsed as if read
from the input stream, allowing for a simplified version of cascaded
parsing~\cite{cascaded-parsing}.  This powerful facility allows several
rules and actions to parse a single input, potentially simplifying both
rules and actions.  XXX EXPAND WITH CHEMICAL FORMULAE EXAMPLE\@.

Glucose: $C_{6}H_{12}O_{6}$
Hydrocarbon (polymer): $CH_{3}{(CH_{2})}_{50}CH_{3}$

>>>>

Each action is a separate procedure written to process a particular
category of input, e.g.\ rejecting a delivery attempt.  There may be many
input variants within one input category; in general each action will
handle one category of inputs, with each rule recognising one input
variant.  It is anticipated that parsers based on this architecture will
have a high ratio of rules to actions, with the aim of having simple rules
and tightly focused actions.  An action may need to process different input
variants in slightly different ways, but large changes in processing
indicate the need for a new action and a new category of input; if an
action becomes overly complicated it starts to turn into a monolithic
parser, with too much logic contained in a single procedure.

The ability to easily add special purpose actions to deal with difficulties
and new requirements that are discovered during parser development is one
of the strengths of this architecture.  Instead of writing a single
monolithic function that processes every input and must be modified to
support new behaviour, with all the attendant risks of adversely affecting
the existing parsing, when a new requirement arises an independent action
can be written to satisfy it.  Sometimes the new action will require the
cooperation of other actions, e.g.\ to set or check a flag, so actions are
not always self-contained, but there will still be a far lower degree of
coupling and interdependency than in a monolithic parser.

During development of the Postfix log parser it became apparent that in
addition to the obvious variety in log~lines there were many complications
to be overcome.  Some were the result of deficiencies in Postfix's logging,
and some of those were rectified by later versions of Postfix, e.g.\
identifying bounce notifications
(\sectionref{identifying-bounce-notifications}); others were due to the
vagaries of process scheduling, client behaviour, and administrative
actions.  All were successfully accommodated in the Postfix log parser:
adding new actions was enough to overcome several of the complications;
others required modifications to a single existing action to work around a
difficulty; the remainder were resolved by adapting existing actions to
cooperate and exchange extra data (via the framework), changing their
behaviour as appropriate based on that extra data.  Every architecture
should aim to make the hard things possible as well as making the easy
things easy; the successful implementation of the Postfix log parser
demonstrates that this architecture achieves that aim.

@@

Actions may return a modified input line to the framework that will be
parsed as if read from the input stream, allowing for a simplified version
of cascaded parsing~\cite{cascaded-parsing}.  This powerful facility allows
several rules and actions to parse a single input, potentially simplifying
both rules and actions.  XXX EXPAND WITH CHEMICAL FORMULAE EXAMPLE\@.

A simple example is to have one rule and action removing comments from
input lines, so that other rules and actions do not have to handle comments
at all; obviously if comment characters can be escaped or embedded in
quoted strings the implementation needs to be more complex, but for some
inputs this can greatly simplify subsequent parsing.

XXX IS WEIGHT THE CORRECT WORD HERE\@?  SHOULD IT BE xxx WEIGHT\@?  GET
GREG TO PROOFREAD THIS SECTION\@.

A more complicated example is determining the weight of a compound given
its chemical formula.  A chemical formula is a sequence of one or two
character chemical element abbreviations, each optionally followed by a
number indicating how many of that element are present in the formula (if
omitted the number is assumed to be one).  Some parts of a formula may be
repeated: they are enclosed in parentheses, and the closing parenthesis is
followed by a number indicating how many are present in the formula.  A
monomer is a simple compound, i.e.\ it does not have any repeated portions;
compounds with repeated portions are called polymers.  E.g.\ glucose (a
monomer) has the chemical formula $C_{6}H_{12}O_{6}$ and hydrocarbon (a
polymer) is $CH_{3}{(CH_{2})}_{50}CH_{3}$.  The atomic weight of an
individual element is given in periodic tables, but one element in
particular will be important in the later explanation: Hydrogen,
abbreviated H, has an atomic weight of 1.  The weight of a given compound
is the sum of the atomic weights of each of the elements multiplied by the
number of times they occur; this is simple to calculate for a monomer, but
it is harder for a polymer because the weight of the repeated monomer must
be multiplied by the number of times the monomer is present in the compound
(e.g.\ for hydrocarbon the weight of $CH_{2}$ must be multiplied by $50$).
Calculating the weight of a monomer requires a single rule and action: the
action separates each element in the monomer, looks up its weight,
multiplies it by the number of times it is present (one if unspecified),
and adds up the results.  The rule's regex is \verb!^([A-Z]([a-z])?\d*)+$!,
which matches:

    \item The start of the input.

    \item One or more of: an uppercase character, optionally followed by a
        lowercase character, followed by zero or more digits.

    \item The end of the input.

The regex does not guarantee that the chemical element abbreviations in the
formula are correct; the regex could be modified to check that by replacing
\verb![A-Z]([a-z])?! with all the valid abbreviations, simplifying the
action at the cost of complicating the regex (which could be automatically
generated).

Parsing a polymer is more difficult because of the repeated monomers: each
monomer's weight needs to be calculated, multiplied by the number of times
it is present in the compound, and that figure added to the sum.  XXX CAN
POLYMERS BE NESTED\@?  Without cascaded parsing the action would need to
maintain state when recursively parsing the monomers embedded in the
polymer.  With cascaded parsing available three rules and actions could be
used as shown in \tableref{Regexes and Actions to calculate the weight of a
chemical compound}:

\newlength{\regexwidth}
% This figure from experimentation; set the length to 0pt, see how big a
% hbox latex complains about, and then set the length.  Stupid latex cannot
% do \settowidth{\verb}.  Poo.
\setlength{\regexwidth}{130pt}
\newlength{\actionwidth}
\setlength{\actionwidth}{\textwidth}
\addtolength{\actionwidth}{-\regexwidth}
% This is the width of the number column, roughly.
\addtolength{\actionwidth}{-30pt}
% This figure also from experimentation.
\addtolength{\actionwidth}{-24pt}
\begin{table}[ht]
    \caption{Regexes and Actions to calculate the weight of a chemical
    compound}
    \empty{}\label{Regexes and Actions to calculate the weight of a
    chemical compound}
    \begin{tabular}{lp{\regexwidth}p{\actionwidth}}

        \tabletopline{}%
        No.   & Regex                       & Action \\
        \tablemiddleline{}%
        1     & \verb'((?!H\d+([()]|$))'    & Parse a monomer as described above,  \\
              & \verb' ([A-Z]([a-z])?\d*)+)'& returning \verb!Hnnn!, where \verb!nnn! is
                                              the weight of the monomer.  The first part
                                              of the regex prevents it from parsing what
                                              it returns (\verb!Hnnn!) and the parsing
                                              process getting stuck in a loop.  \\
        2     & \verb!\(H(\d+)\)(\d+)!      & Parse a repeated monomer that has been
                                              rewritten by the first rule and action,
                                              returning \verb!Hnnn!, where \verb!nnn! is
                                              the product of the first and second numbers
                                              captured by the regex.  \\
        3     & \verb!^H(\d+)+$!            & Repeated applications of rules and
                                              actions one and two will have rewritten the
                                              input to a sequence of one or more
                                              \verb!Hnnn!, where \verb!nnn! is the weight
                                              of the compound.  This action should do
                                              whatever is required with the calculated
                                              weight.  \\
        \tablebottomline{}%

    \end{tabular}
\end{table}

Applying the rules and actions above to the hydrocarbon formula results in
this sequence:

\begin{tabular}{lrl}

    1 & Start with              & $CH_{3}{(CH_{2})}_{50}CH_{3}$ \\
    2 & Apply rule and action 1 & $H_{15}{(CH_{2})}_{50}CH_{3}$ \\
    3 & Apply rule and action 1 & $H_{15}{(H_{14})}_{50}CH_{3}$ \\
    3 & Apply rule and action 1 & $H_{15}{(H_{14})}_{50}H_{15}$ \\
    5 & Apply rule and action 2 & $H_{15}H_{700}H_{15}$ \\
    6 & Apply rule and action 1 & $H_{730}$ \\
    7 & Apply rule and action 3 & 730 \\

XXX ARE THERE ANY PROBLEMS TO BE ADDRESSED HERE\@?  I DO NOT THINK THAT I
NEED TO WORRY ABOUT ORDERING OR PRIORITY, BUT I SHOULD REVIEW TO BE SURE\@.
